Basics of Hardware Cache Optimization

### 2017-05-26 tex/types.tex
Add amazing table and tweak things

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
   \inputminted[firstline=12]{c}{access-times/access-times.c}
   \captionof{listing}{}
   % \captionof{listing}{Random Reads\label{lst:access-times}}
   \label{lst:access-times}
\end{center}

@@

\Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
when not having defined \mintinline{text}{BASELINE}.  That is, only the cycles used by the
main loop are counted, not those for initialization.
I divided by \mintinline{text}{N} to get the cycles spend \alts{per, on each} loop
iteration.

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{TODO: table of actual data in appendix?}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets evicted.  The
first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.

The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
increase when the array size matches that of the \gls{l2} is more striking than for the
\gls{l1d} before; possibly because \gls{l2} is a unified cache that also holds
instructions.  Eventually, more and more accesses go to main memory, \alts{causing,
incurring} delays of up to 200 cycles.

TODO~\cite[cf.][17]{drepper2007}.

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={2048 as 2, 8192 as 8, 32768 as \textbf{32},
                                 131072 as 128, 524288 as \textbf{512}, 2097152 as 2048,
                                 8388608 as 8192, 33554432 as 32768, 134217728 as
                                 % $128\cdot 2^{10}$}}},
                                 % $2^{17}$}}},
                                 131072}}},
               grid={at={32768, 524288}},
               label={Array Size (KiB)}, length=0.8\textwidth},
       y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
       % y axis={logarithmic,
       %         ticks={major={at={3, 15, 23, 190}}},
       %         grid={at={15, 23}},
       %         label={Cycles / Element}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=access-times/access-times.csv, separator=\space];
   \end{tikzpicture}
   % \caption{Sequential Read Access}
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}

>>>>

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
   \inputminted[firstline=12]{c}{access-times/access-times.c}
   % \captionof{listing}{}
   % \captionof{listing}{Randomly Access Elements of \mintinline{text}{array}}
   \captionof{listing}{Randomly Read Array Elements}
   % \captionof{listing}{Random Reads\label{lst:access-times}}
   \label{lst:access-times}
\end{center}

@@

% \Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
% when not having defined \mintinline{text}{BASELINE}.
\Cref{fig:access-times} shows the extra CPU cycles used by \cref{lst:access-times}
\alts[2]{above, in addition to, compared to} the \mintinline{text}{BASELINE} version for
different array sizes.
That is, only the cycles used by the main loop are \alts{given, counted}, not those for
initialization.  I divided by \mintinline{text}{N} to get the cycles spent \alts{per, on
each} loop iteration.

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{The numerical results are shown in \cref{tab:access-times}.}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets evicted.  The
first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.

The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
increase when the array size matches that of the \gls{l2}, 512 KiB, is \alts{greater, more
striking} than for the \gls{l1d} before; possibly because \gls{l2} is a unified cache that
also holds instructions.  Eventually, more and more accesses go to main memory,
\alts{causing, incurring} delays of up to 200 cycles~\cite[cf.][17]{drepper2007}.

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization[%
         scientific axes=clean,
         x axis={%
            logarithmic,
            ticks={major={
               /pgf/number format/int detect,
               at={
                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
                  8388608 as \pgfmathprintnumber{8192},
                  33554432 as \pgfmathprintnumber{32768},
                  134217728 as \pgfmathprintnumber{131072},
                  % $128\cdot 2^{10}$}}},
                  % $2^{17}$}}},
            }}},
            grid={at={32768, 524288}},
            label={Array Size (KiB)},
            length=0.8\textwidth},
         y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
         visualize as scatter,
         scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=access-times/access-times.csv, separator=\space];
   \end{tikzpicture}
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}
